---
title: 'Reproducible Research: Project 1'
author: "YLL"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Dataset: Daily Steps
The dataset for this project is downloaded from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip), 
and unzips to a CSV file.<br>

The file is read into R using read.csv and the NAs are removed for the initial
analyses.<br>

```{r read_data}
data_raw <- read.csv("activity.csv")
data <- data_raw[!is.na(data_raw$steps),]
```

## Average Daily Steps
Using the data with NAs removed, the daily total steps are calculated and 
displayed in a histogram.<br>
The mean and median of the daily steps are computed using R functions mean and
median.<br>

```{r steps_histogram}
daily_steps <- tapply (data$steps, as.factor(data$date), sum)
daily_steps <- daily_steps[!is.na(daily_steps)]
hist(daily_steps)
print(paste("Mean daily steps: ", mean(daily_steps)))
print(paste("Median of daily steps: ", median(daily_steps)))
```

## Average Daily Activity Pattern
To look at the mean of the number of steps every day for each time interval, the
data with NAs removed is used to create a plot.<br>
The interval with the highest number of steps is calculated by identifying the
largest average number of steps across all intervals.<br>

```{r steps_interval_plot}
intervals <- unique(data$interval)
interval_steps <- tapply (data$steps, as.factor(data$interval), mean)
interval_steps <- bind_cols(interval_steps,intervals)
names(interval_steps) <- c("Steps", "Interval")
plot(interval_steps$Interval,interval_steps$Steps,type="l")
max_point <- interval_steps[interval_steps$Steps==max(interval_steps$Steps),]
print (paste("Interval with maximum steps: ", max_point[2]))
```

## Imputing Missing Values
All the NAs from the original data is replaced with the average value of the
time interval to which the missing value belongs.<br>

```{r missing_values}
data_complete <- data_raw
nas <- is.na(data_complete$steps)
num_nas <- sum(nas)
for (i in 1:length(data_complete$steps)) {
  if (is.na(data_complete[i,1])) {
    data_complete[i,1] <- interval_steps[interval_steps$Interval==data_complete[i,3],1]
  }
}
```

## Recalculate Average Daily Steps 
Using the new completed data, the average daily steps is shown in a histogram
and the mean and median are re-calculated.

```{r steps_histogram_complete}
daily_steps_complete <- tapply (data_complete$steps, as.factor(data_complete$date), sum)
daily_steps_complete <- daily_steps_complete[!is.na(daily_steps_complete)]
hist(daily_steps_complete)
print(paste("Mean daily steps: ", mean(daily_steps_complete)))
print(paste("Median of daily steps: ", median(daily_steps_complete)))
```

## Activity Pattern Difference
To explore the difference in activity patterns between weekdays and weekends, 
the completed data is analyzed to produce a plot with separate points for
weekdays and weekends.<br>

First, the "date" column is converted to class Date so that each date can be
classified as either weekday or weekend. A new column indicating whether an
observation is a weekday or not is added to the data.<br>

```{r weekdays1}
data_complete$date <- as.Date(data_complete$date, format="%Y-%m-%d")
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
data_complete <- mutate (data_complete, 
                         isWeekday=(weekdays(data_complete$date) %in% weekdays))
```

Next, two subsets of the data are created - one for the weekdays and another
for the weekends. The average steps per each time interval is calculated for 
each data subset independently. Lastly, both data subset are shown in the same
plot.<br>

```{r weekdays2}
weekday_steps <- filter(data_complete, data_complete$isWeekday==TRUE)
weekend_steps <- filter(data_complete, data_complete$isWeekday==FALSE)

weekday_ave <- tapply (weekday_steps$steps, as.factor(weekday_steps$interval), mean)
weekday_ave <- bind_cols(weekday_ave,intervals)
names(weekday_ave) <- c("Steps", "Interval")

weekend_ave <- tapply (weekend_steps$steps, as.factor(weekend_steps$interval), mean)
weekend_ave <- bind_cols(weekend_ave,intervals)
names(weekend_ave) <- c("Steps", "Interval")

plot(weekday_ave$Interval, weekday_ave$Steps,
     type="l",
     col="blue",
     xlab="Interval", ylab="Steps")
points(weekend_ave$Interval, weekend_ave$Steps, 
       type="l",
       col="red")
legend ("topright", 
        lty=1, 
        col=c("blue","red"),
        legend=c("weekday","weekend"))
```

End of report.